Wait-free Queues with Polygarithmic Step Complexity

### 2022-08-10 article/related.tex
add related, bug fix

\documentclass[12pt]{article}
\usepackage[a4paper]{geometry}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{tikz-qtree}

\renewcommand{\baselinestretch}{1.5}

@@

\title{Wait-free Queues with Polylogarithmic\\ Step Complexity}
\author{Hossein Naderibeni \\supervised by Eric Ruppert}
\maketitle

%\begin{enumerate}
%  \item Importance of implementing shared data structures
%\begin{itemize}
%  \item how it helps developers
%  \item why lock free data structures are better than using locks?
%\end{itemize} 
%  \item  What properties matter for a distributed data structure?
%\begin{itemize}
%  \item Correctness condition: Linearizability definition
%  \item Wait-freeness
%\end{itemize} 
%  \item Previous works on distributed queues: more detailed
%\begin{itemize}
%  \item why we are studying queues
%  \item $p$ factor in time complexity of all previous algorithms
%  \item types of current queues in the literature
%  \item previous impractical universal constructions that use long words
%\end{itemize}
%  \item Our contribution
%\begin{itemize}
%  \item polylogarithmic algorithm with respect to $p$ and $Q$ size of the queue
%\end{itemize}
%  \item Explaining ideas
%\begin{itemize}
%  \item using tournament tree to agree on the linearization order
%  \item double refresh
%  \item how to compute a dequeue response
%  \item how to be sure the number of blocks in the root is related to the size of the queue
%  \item memory management as a goal
%\end{itemize}
%\end{enumerate}

\pagebreak

\section{Introduction}

In this work, we are going to introduce a novel lock-free queue. Linearizable lock-free data structures are the current standard requirements for designing distributed algorithms. All lock-free queues in the literature yet have a common problem in their worst case. Our algorithm solves this problem, performing better than them.

Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. This is why distributed algorithms have become important. It is not hard to see why multiple processes cannot update sequential data structures designed for one process. For example, consider two processes trying to insert some values into a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks it, and others cannot use it until the lock is released. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress. Current approach is to create a lock-free queue data structure that developers can safely use without the disadvantages of locks.

The question that may arise is, ``What properties matter for a lock-free data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

The standard safety condition is called \textit{linearizability}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response. Figure \ref{fig::goodexample} is an example of an execution on a linearizable  queue that is initially empty. The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation. Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}. The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.

\begin{figure}[t]
  \center\includegraphics[scale=0.5]{pics/good}
  \caption{\label{fig::goodexample}An example of a linearizable execution. Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
\end{figure}

\begin{figure}[t]
  \center\includegraphics[scale=0.5]{pics/bad}
  \caption{\label{fig::badexample}An example of an execution that is not linearizable. Since \texttt{Enqueue(A)} has completed before \texttt{Enqueue(B)} is invoked the \texttt{Dequeue()} should return \texttt{A} or nothing.}
\end{figure}

There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom. An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps. We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates. A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm there might be an operation that takes infinitely many steps but never terminates.
%From now on, we are going to focus on wait freedom.

%Since executions on a shared data structure are different from sequential ones, the correctness condition also differs. The strongest correctness condition is linearizability, defined as "...". Lock-free is good, but a wait-free algorithm guarantees us every operation terminates after finite steps.

@@

\begin{figure}[t]
  \center\includegraphics[scale=0.5]{pics/msqueue}
  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. In the first diagram the first element has been dequeued. Red arrows show new pointers and gray dashed arrows show the old pointers.}
\end{figure}

@@

\begin{figure}[t]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption{Baskets queue idea. There is a time that all operations in a basket were running concurrently, but only one has succeeded to do \texttt{CAS}. To order the operations in a basket, the mechanism in the algorithm for processes is to \texttt{CAS} again. The successful process will be the next one in the basket and so on.}
\end{figure}

@@

\begin{figure}[t]
  \center\includegraphics[scale=0.5]{pics/sundell}
    \caption{\label{fig::sundell}Global pointers point to arrays. Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
\end{figure}

@@

\subsection{Universal constructions}
Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}. A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also introduced a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}. His universal construction can be used to create any data structure, but its implementation is not practical because of using unreasonably large-sized \texttt{CAS} operations.

@@

%There is a connection between queues and universal constructions. We can implement a universal construction using a queue. We can store operations in the shared queue and compute operations' responses using the queue's content. Some impractical universal constructions are using big words as "...".

\section{Our Contribution}
In this work, we design a queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the size of the queue, avoiding the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays. We introduce a data structure that allows processes to agree on the linearization ordering of their operations using $\textsc{O}(\log p)$ \texttt{CAS} per operation called a \textit{block tree}. Then we use the block tree as a stepping stone towards our queue algorithm.
A block tree is a tournament tree shared among $p$ processes (see Figure~\ref{fig::blocktree}). Each process has a leaf, and it appends its operations to its leaf. After that, the process tries to propagate its new operation up to the tree's root. An ordering of operations propagated up to a node is stored in that node. All processes agree on the sequence stored in the root and this is used as the linearization ordering. Our idea is similar to Jayanti and Petrovic's multi-enqueuer single-dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent dequeues.

\begin{figure}[t]
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption{ّ\label{fig::blocktree}In the block tree each process has a leaf and in each node there is an ordering of operations stored. Each node tries to propagate its operations up to the root, which stores the final ordering of all operations.}  
\end{figure}

The goal here is to ensure that in each propagate step the new operations are propagated up to the parent in $O(\log p)$ steps (see~Figure~\ref{fig::propagstep}). Then, a dequeue operation uses the linearization ordering to compute its answer.

In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using a \texttt{CAS} operation. The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in $n$'s children before the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
%Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$), which has appended the talked operations.
Instead of storing operations explicitly in the nodes, we only keep track of the number of them. This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$). To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \texttt{Refresh}() procedure and only propagate the block of the new operations.

\begin{figure}[t]
\begin{center}
\begin{subfigure}[b]{.49\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c||c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c||c|c|}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
\end{tikzpicture}}
  \caption{Operations after $||$ are new.}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.49\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ & $op_{l_4}$ & $op_{l_5}$ & $op_{r_3}$ & $op_{r_4}$  \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c|c|c||}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c|c||}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
\end{tikzpicture}}
  \caption{New operations are added to the parent node.}
\end{subfigure}
\caption{\label{fig::propagstep} Successful \texttt{Refresh}, operations in children after $||$ are new.}
\end{center}
\end{figure}

We also implement methods \texttt{Get(i)}, \texttt{Index(op)} to get the $i$th propagated operation and compute the rank of a propagated operation in the linearization. \texttt{Get(i)} finds the block containing the $i$th operation in the root and then finds its sub-block recursively to reach a leaf. \texttt{Index()} is  similar but more complicated, finding super-blocks from a leaf to the root. The main challenge in each level of \texttt{Get(i)} and \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$. After appending operations to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}. Now we can implement Enqueue and Dequeue using our block tree. An \texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree. To do a \texttt{Dequeue()}, process $p$ first appends a \texttt{DEQ} operation to the tree. Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}. Finally $p$ returns the argument of that \texttt{ENQ} using \texttt{Get(i)}.

\texttt{Get()} and \texttt{Index()} search among blocks in each level of the tree to find the sub-block or super-block containing the given operation. Each block stores a constant amount of information (like prefix sums) to allow binary searches to find the required block in a node quickly.

The algorithm works as a queue, but \texttt{Get(i)}  may take a long time since it has to find the block containing the $i$th operation at the root level. We came up with a garbage collection only for the root level to reduce the time to find this block to $O(\log q)$. The main idea is to remove a block after its operations are no longer needed. We will store the blocks in the root in a persistent red-black tree to be able only to maintain the remaining information of the queue. It is possible to search in persistent red-black trees with size $n$ in $O(\log n)$ steps and split the tree by a key in $O(\log n)$ steps.  Finally, the algorithm works in $O(\log^2 p +\log q)$ steps.

\section{Next steps}
The current algorithm only works as a queue, but other data structures like stacks share similarities. It might be possible to come up with a generalization for all these types. We have not yet considered space usage, and we use infinite-sized arrays that are not practical. Our future work is to manage the memory of the tree and make it practical.

\section{Planning}
The plan is to finish the writing of the thesis by August 2021. The thesis will contain:
\begin{itemize}
  \item More detailed discussion of previous work
  \item Description of the algorithm, layer by layer
  \item Pseudocode of the algorithm
  \item Proof of linearizability and time analysis
\end{itemize}

We expect to submit a paper of our work to \textit{36th International Symposium on Distributed Computing
October 25-27, 2022, Augusta, Georgia, USA} by May 13.

\bibliography{main} 
\bibliographystyle{ieeetr}

>>>>

\documentclass[12pt]{article}
\renewcommand{\baselinestretch}{1.8}

\usepackage[b4paper]{geometry}
\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{showkeys}
\usepackage{amsmath}

\usepackage{multicol}
\setlength\columnsep{24pt}

\algnewcommand\algorithmicforeach{\bf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}

\algdef{S}[FUNCTION]{Function}
   [3]{{\tt{\sl{#1}}} \textproc{\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}

\algdef{E}[FUNCTION]{EndFunction}
   [1]{\algorithmicend\ \tt{\textproc{#1}}}

\algrenewcommand\Call[2]{\tt{\textproc{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}

\newcommand\keywordfont{\sffamily\bfseries}
\algrenewcommand\algorithmicend{{\keywordfont end}}
\algrenewcommand\algorithmicfor{{\keywordfont for}}
\algrenewcommand\algorithmicforeach{{\keywordfont for each}}
\algrenewcommand\algorithmicdo{{\keywordfont do}}
\algrenewcommand\algorithmicuntil{{\keywordfont until}}
\algrenewcommand\algorithmicfunction{{\keywordfont function}}
\algrenewcommand\algorithmicif{{\keywordfont if}}
\algrenewcommand\algorithmicthen{{\keywordfont then}}
\algrenewcommand\algorithmicelse{{\keywordfont else}}
\algrenewcommand\algorithmicreturn{{\keywordfont return}}

\renewcommand\thealgorithm{}
\newcommand{\setalglineno}[1]{%
  \setcounter{ALC@line}{\numexpr#1-1}}

\newcommand{\sub}[1]{\textsubscript{#1}}
\renewcommand{\tt}[1]{\texttt{#1}}
\renewcommand{\sl}[1]{\textsl{#1}}
\renewcommand{\it}[1]{\textit{#1}}
\renewcommand{\sc}[1]{\textsc{#1}}
\renewcommand{\bf}[1]{\textbf{#1}}
\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
\newcommand{\cmt}[1]{\Comment{#1}}
\newcommand{\head}{head}
\newcommand{\size}{size }

\usepackage{amsmath,amssymb,amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{observation}[theorem]{Observation}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{invariant}[theorem]{Invariant}

@@

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.4]{pics/msqueue}
  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. In the first diagram the first element has been dequeued. Red arrows show new pointers and gray dashed arrows show the old pointers.}
\end{figure}

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption{Baskets queue idea. There is a time that all operations in a basket were running concurrently, but only one has succeeded to do \texttt{CAS}. To order the operations in a basket, the mechanism in the algorithm for processes is to \texttt{CAS} again. The successful process will be the next one in the basket and so on.}
\end{figure}

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/sundell}
    \caption{\label{fig::sundell}Global pointers point to arrays. Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
\end{figure}

@@

\subsection{Universal Constructions}
Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}. A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also introduced a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}. His universal construction can be used to create any data structure, but its implementation is not practical because of using unreasonably large-sized \texttt{CAS} operations.

@@

%There is a connection between queues and universal constructions. We can implement a universal construction using a queue. We can store operations in the shared queue and compute operations' responses using the queue's content. Some impractical universal constructions are using big words as "...".
\section{Adiitional ones}
\subsection{Tsigas and Zhang}
Tsigas and Zhang \cite{DBLP:conf/spaa/TsigasZ01} present a cyclic array queue that allows the head and tail to lag at most m nodes behind the actual head and tail of the queue, so the amortized number of CAS executions per operation is 1 + 1/m. Their algorithm is limited to bounded queues due to their static allocation.

\subsection{Kogan and Herlihy}
Previous works present concurrent constructs that combine multiple operations into a single operation on the shared object. We chose to combine operations and apply them as batches, to increase scalability. The work of Kogan and Herlihy \cite{DBLP:conf/podc/KoganH14} is the closest to this work. They propose alternative definitions for linearizability of executions with batches, including MF-linearizability, which we use. They describe very simple implementations of stacks, queues, and linked lists that demonstrate the benefits of using futures.

\subsection{Yang and Mellor-Crummey}
Because of the problem of failure of CASes one way is to use other primitive objects other than CAS objects. Yang and Mellor-Crummey \cite{DBLP:conf/ppopp/YangM16} utilize fetch-and-add as well, to form a wait-free queue.  They present the first linearizable and wait-free FIFO queue based on fetch-and-add (FAA). While compare-and-swap (CAS) based non-blocking algorithms may perform poorly due to work wasted by CAS failures, algorithms that coordinate using FAA, which is guaranteed to succeed, can in principle perform better under high contention.

I could not compute its complexity but I guess it is hard or not important

\bibliography{main}
